---
title: Risk factors for mortality in patients with post COVID-19 vaccination adverse reactions 
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Data Prepration 
### 1.1 Load the data 
```{r}
setwd("~/covid-adv")
covid_adv <-read.csv("./covid_adv.csv")
print(head(covid_adv, n = 5))
```

### 1.2 Transform the data
#### Label Encoding 
```{r}
# reorder columns 
df <-covid_adv[c("VAERS_ID","AGE_YRS","AGE_GRP","SEX","DIED","HOSPITAL",
                 "DISABLE","L_THREAT","BIRTH_DEFECT","Asthma_c","Hypertension_c",
                "Diabetes_c","Anxiety_c","Allergy_c","Arthritis_c","Depression_c",
                      "Hypothyroidsm_c","High_Cholesterol_c","Heart_Disease_c","GERD_c",
                      "Cancer_c","Obesity_c","Migraine_c","Kidney_Disease_c",
                      "COVID19_Positive","Headache","Pyrexia","Fatigue","Chills",
                      "Swelling","Rash","Dizziness","Nausea","Pruritus","Erythema",
                      "Myalgia","Dyspnoea","Paraesthesia", "Hypoaesthesia", "Asthenia",
                      "Muscle_weakness","Vomiting","Hyperhidrosis","Urticaria",
                      "Diarrhoea","Abdominal","Malaise","Tinnitus","Palpitations",
                      "Flushing","Blood_pressure_increased","Feeling_abnormal",
                      "Feeling_hot","Tremor","Lymphadenopathy","Syncope",
                      "Throat_irritation","Unresponsive_stimuli","Burning_sensation",
                      "Muscle_tightness","Skin_issues","Dyskinesia",
                      "Heart_rate_increased","Heart_rate_decreased",
                      "Heart_rate_irregular","Paralysis","Tachycardia",
                      "Anaphylactic_reaction","Cardiac_event","Chest_pain","Pain",
                      "Mobility_decreased","Stroke","Influenza","SARS.COV.2",
                      "Appetite_decreased","Induration","Sleep.disorder",
                      "Condition_aggravated", "Joint_issues","VAX_MANU")]


df[['SEX']] <-ifelse(df[['SEX']] == 'F', 1, 0)
df[['DIED']] <- ifelse(df[['DIED']] == 'Yes', 1, 0)
cat_cols <- colnames(covid_adv)
cols_to_remove <-c('VAERS_ID', 'AGE_YRS', 'DIED',
                    'AGE_GRP','SEX', 'VAX_MANU')
cols <- cat_cols [! cat_cols %in% cols_to_remove]

for (c in cols){
  df[[c]] <- ifelse(df[[c]] =='Yes', 1, 0)
}

# set VAX_MANU as factor variable
df$VAX_MANU <-factor(df$VAX_MANU)
df$AGE_GRP <- factor(df$AGE_GRP)
```

## 2. Statistical Analysis

### 2.1 Chi-Squared Test
Chi-squared test was used to assess if there is an **association** between

+ patient's demographic factor (age group, gender)
+ patient's medical history (comorbidities)
+ having a particular adverse reaction symptom
+ other factor (e.g, manufacturer of vaccination)

and mortality. 

```{r, fig.width = 5, fig.height = 5}

library(MASS)
library(ggplot2)
chi_func <-function(col_name){
  c_table <-table(df$DIED, col_name)
  c_test <-chisq.test(c_table)
  return(c_test$p.value)
}

#cols to test
test_cols <- colnames(df)
cols_to_remove <-c('VAERS_ID', 'AGE_YRS', 'DIED')
test_cols <- test_cols [! test_cols %in% cols_to_remove]

#loop for applying chi-squared test
p_values <- c() 
for(i in 1:length(test_cols)){
  test_col <- test_cols[i]
  p <- chi_func(df[[test_col]])
  p_values <- append(p_values, p)
}

# data frame for p-value and corresponding variable
chisq_test_result <- data.frame(test_cols, p_values)
colnames(chisq_test_result) <- c('Variable', 'p_values')

# sort out variables that are associated with mortality with p < 0.05
significant <-c()
for(j in chisq_test_result[['p_values']]){
  if(j <0.05){
    significant <-append(significant, 'Yes')
  }else{
    significant <-append(significant, 'No')
  }
}
chisq_test_result$Significant <- significant

# -log(p-value)
log_10 <- c()
for(p in chisq_test_result[['p_values']]){
  l <- -log10(p)
  log_10 <- append(log_10, l)
}

chisq_test_result$LogP_value <- log_10

print(head(chisq_test_result, n = 5))


# chi-squared test results visualisation
ggplot(data = chisq_test_result, aes(x = reorder(Variable, LogP_value), y = LogP_value )) + geom_bar(stat = 'Identity', fill = 'steelblue') + coord_flip() + labs(title = 'Chi-squared test results', y = " -Log10 p-value", x = "Features") + geom_hline(yintercept = -log10(0.05),color = 'gold1', size = 0.5) + coord_flip(ylim=c(0, 10)) 
```

### 2.2 Logistic Regression Analysis

#### 2.2.1 Adjusting for age and sex

Logistic regression analyses were performed to assess the respective independent effects of symptoms and other factors on mortality after adjusting for age and gender.

- Age
```{r age}
# reference set to alive 
df$DIED <- factor(df$DIED, levels = c(0, 1), labels = c('No', 'Yes'))
df$DIED<- relevel(df$DIED, ref = 'No')

# age 
##age in yrs
OR <- c()
P <-c()
CI_low <- c()
CI_high <- c()
variable_name <-c('Age in years', '+80', '18 - 34', '35 - 49', '50 - 64', 
                  '65 - 79')

age_m <- glm(DIED ~ AGE_YRS, family = binomial, data = df)
or <- coef(age_m)[2] # coefficient
OR <- append(OR, exp(or))
p  <- coef(summary(age_m))[, 'Pr(>|z|)'][2]
P <- append(P, p)
conf_age <- confint(age_m)
conf_low <- conf_age[2,1] # low CI
conf_high <- conf_age[2,2] # high CI
CI_low <- append(CI_low, exp(conf_low))
CI_high<- append(CI_high, exp(conf_high))

# age in grp
df$AGE_GRP<- relevel(df$AGE_GRP, ref = '0-17')
age_m_grp <- glm(DIED ~ AGE_GRP, family = binomial, data = df)
conf_age_grp <- confint(age_m_grp)
for(i in 2:6){
  or <- coef(age_m_grp)[i] # coefficient
  OR <- append(OR, exp(or))
  p  <- coef(summary(age_m_grp))[, 'Pr(>|z|)'][i]
  P <- append(P, p)
  conf_low <- conf_age_grp[i,1] # low CI
  CI_low <- append(CI_low, exp(conf_low))
  conf_high <- conf_age_grp[i,2] # high CI
  CI_high<- append(CI_high, exp(conf_high))
}

## create df
age_df <- data.frame(variable_name, OR, CI_low, CI_high, P)
colnames(age_df) <- c('Factors', 'OR','CI_2.5', 'CI_97.5', 'P_value')

## reorder 
library(dplyr)
age_order <- c('Age in years', '18 - 34', '35 - 49', 
               '50 - 64', '65 - 79', '≥80')
age_df <- age_df %>%
  slice(match(age_order, Factors))
```

- Age & Sex
```{r age & sex}

# age & sex
age_sex_m <- glm(DIED ~ AGE_YRS + SEX , family = binomial, data = df)
or <- exp(coef(age_sex_m)[3]) # coefficient
p  <- coef(summary(age_sex_m))[, 'Pr(>|z|)'][3]
conf_age_sex <- confint(age_sex_m)
conf_low <- exp(conf_age_sex[3,1]) # low CI
conf_high <- exp(conf_age_sex[3,2]) # high CI

sex_df <- data.frame('Gender, Female vs Male', or, conf_low, conf_high, p)
names(sex_df) <- colnames(age_df)
age_sex_df <- rbind(age_df,sex_df )

```
- Other variables

```{r other variables}
# other variables 
features <- colnames(df)
feature_to_remove <- c('VAERS_ID', 'AGE_YRS', 'AGE_GRP', 'DIED', 'SEX', 'VAX_MANU')
features <- features [! features %in% feature_to_remove]

OR <- c()
P <-c()
CI_low <- c()
CI_high <- c()
factor_names <- colnames(df[6:80])

for(f in seq_along(features)){
  vars <-as.formula(sprintf("DIED ~ AGE_YRS + SEX + %s", features[f]))
  logit <-glm(formula = vars, family = binomial, data = df)
  or <- coef(logit)[4] # coefficient
  OR <- append(OR, exp(or)) # odd ratio
  P <- append(P, coef(summary(logit))[, 'Pr(>|z|)'][4])
  conf_other <- confint(logit)
  conf_low <- conf_other[4,1] # low CI
  conf_high <- conf_other[4,2] # high CI
  CI_low <- append(CI_low, exp(conf_low))
  CI_high<- append(CI_high, exp(conf_high))
}

other_df <- data.frame(factor_names, OR, CI_low, CI_high, P)
colnames(other_df) <- c('Factors', 'OR','CI_2.5', 'CI_97.5', 'P_value')

# combine dfs
odd_df <- rbind(age_sex_df,other_df )
```

```{r vaccine manufacturer}
# vaccine manufacturer
vac_m <- glm(DIED ~ AGE_YRS + SEX + VAX_MANU, family = binomial,
             data = df)
or <- exp(coef(vac_m)[4]) # coefficient
p  <- coef(summary(vac_m))[, 'Pr(>|z|)'][4]
conf_vac <- confint(vac_m)
conf_low <- exp(conf_vac[4,1]) # low CI
conf_high <- exp(conf_vac[4,2]) # high CI

MODERNA <- data.frame('MODERNA', or, conf_low, conf_high, p)
names(MODERNA) <-colnames(odd_df)
odd_df <- rbind(odd_df, MODERNA)

or <- exp(coef(vac_m)[5]) # coefficient
p  <- coef(summary(vac_m))[, 'Pr(>|z|)'][5]
conf_low <- exp(conf_vac[5,1])# low CI
conf_high <- exp(conf_vac[5,2]) # high CI

PFIZER <- data.frame('PFIZER', or, conf_low, conf_high, p)
names(PFIZER) <-colnames(odd_df)
odd_df <- rbind(odd_df, PFIZER)

write.csv(odd_df, 'logit_results_uni.csv')

```


#### 2.2.2 Fully adjusted 

```{r}
library(R.utils)

full_m <- glm(DIED ~ . -VAERS_ID -AGE_GRP, family = binomial, data = df)

OR <- c()
P <- coef(summary(full_m))[, 'Pr(>|z|)'][2:79]
P <- insert(P, 61, values = 999) # need to mark as NA later
CI_low <- c()
CI_high <- c()

conf <- confint(full_m)


for(j in 2:80){
  or <- coef(full_m)[j]
  OR <- append(OR, exp(or))
  conf_low <- conf[j,1]
  CI_low <- append(CI_low, exp(conf_low))
  conf_high <- conf[j,2]
  CI_high<- append(CI_high, exp(conf_high))
}

"# insert values empty values for age group 
OR <- insert(OR, ats = 2:10, values =rep(" ", 9))
P <- insert(P, ats = 2:10, values =rep(" ", 9))
CI_low <- insert(CI_low, ats = 2:10, values =rep(" ", 9))
CI_high <- insert(CI_high, ats = 2:10, values =rep(" ", 9))"

full_df <- data.frame(OR, CI_low, CI_high, P)
names(full_df) <- c('OR','CI_2.5', 'CI_97.5', 'P_value')

write.csv(full_df, 'logit_results_full.csv')

```

#### 2.2.3 Visualisation (adjusted for age & sex)
```{r fig.width = 5, fig.height = 5}
# visualisation of logistic regression results
boxLables <- odd_df$Factors

ggplot(odd_df, aes(x = OR, y = length(odd_df$Factors):1)) + 
  geom_vline(aes(xintercept = 1), size = .40, linetype = 'dashed',
             color = 'red') + 
  geom_errorbarh(aes(xmax = CI_97.5, xmin = CI_2.5), size = .5,
                 height = .2, color = 'gray50') +
  geom_point(size = 2.5, color = 'orange') +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  scale_y_continuous(breaks = length(odd_df$Factors):1, 
                     labels = odd_df$Factors) +
  #scale_x_continuous(breaks = seq(0, 3, 0.5)) +
  scale_x_continuous(limits = c(0, 3, 0.5)) +
  ylab("") + xlab("Odds Ratio")

```
## 3. Tables
### 3.1 Table 1 - Descriptive statistics of baseline characteristics 
```{r}
library(compareGroups)

# reorder columns 
covid_df <- covid_adv[c("VAERS_ID","AGE_YRS","AGE_GRP","SEX","DIED","HOSPITAL",
                      "DISABLE","L_THREAT","BIRTH_DEFECT","Asthma_c","Hypertension_c",
                      "Diabetes_c","Anxiety_c","Allergy_c","Arthritis_c","Depression_c",
                      "Hypothyroidsm_c","High_Cholesterol_c","Heart_Disease_c","GERD_c",
                      "Cancer_c","Obesity_c","Migraine_c","Kidney_Disease_c",
                      "COVID19_Positive","Headache","Pyrexia","Fatigue","Chills",
                      "Swelling","Rash","Dizziness","Nausea","Pruritus","Erythema",
                      "Myalgia","Dyspnoea","Paraesthesia", "Hypoaesthesia", "Asthenia",
                      "Muscle_weakness","Vomiting","Hyperhidrosis","Urticaria",
                      "Diarrhoea","Abdominal","Malaise","Tinnitus","Palpitations",
                      "Flushing","Blood_pressure_increased","Feeling_abnormal",
                      "Feeling_hot","Tremor","Lymphadenopathy","Syncope",
                      "Throat_irritation","Unresponsive_stimuli","Burning_sensation",
                      "Muscle_tightness","Skin_issues","Dyskinesia",
                      "Heart_rate_increased","Heart_rate_decreased",
                      "Heart_rate_irregular","Paralysis","Tachycardia",
                      "Anaphylactic_reaction","Cardiac_event","Chest_pain","Pain",
                      "Mobility_decreased","Stroke","Influenza","SARS.COV.2",
                      "Appetite_decreased","Induration","Sleep.disorder",
                      "Condition_aggravated", "Joint_issues","VAX_MANU")]


# Factor variables 
covid_df$DIED <- factor(covid_df$DIED, levels = c('No', 'Yes'), labels = c('Alive',
                                                                      'Died'))
covid_df$SEX <- factor(covid_df$SEX, levels = c('M', 'F'), labels = c('Male',
                                                                      'Female'))
covid_df$AGE_GRP <-factor(covid_df$AGE_GRP, levels = c('0-17', '18-34', '35-49','50-64', 
                                      '65-79','>= 80'), 
                          labels =  c('0 - 17', '18 - 34', '35 - 49','50 - 64', 
                                      '65 - 79','>= 80'))
#cat_cols
cat_cols <- colnames(covid_df)
cols_to_remove <-c('VAERS_ID', 'AGE_YRS','VAX_MANU')
cat_cols<- cat_cols [! cat_cols %in% cols_to_remove]

for(i in 1:length(cat_cols)){
  cat_col <- cat_cols[i]
  covid_df[[cat_col]] <- factor(covid_df[[cat_col]])
}

# Label variables 
library(stringr)
library(Hmisc)

label(covid_df$SEX) <- 'Gender, n(%)'

comorb_cols <- colnames(covid_df[10:24])
for (j in 1:length(comorb_cols)){
  col <- comorb_cols[j]
  label(covid_df[[col]]) <- str_sub(col, 1, -3)
}

label(covid_df$COVID19_Positive) <- 'Covid-19 positive history'
label(covid_df$VAX_MANU) <- 'Vaccine manufacturer'
label(covid_df$AGE_YRS) <- 'Age, mean(SD)'
label(covid_df$AGE_GRP) <- 'Age by group, n(%)'
label(covid_df$HOSPITAL) <- 'Hospital admission status, n(%)'
label(covid_df$DISABLE) <- 'Post-vaccination disability, n(%)'
label(covid_df$L_THREAT) <- 'Life-threatening event, n(%)'
label(covid_df$BIRTH_DEFECT) <- 'Birth defeat, n(%)'

# computing descriptives 
res<- compareGroups(DIED ~ . - VAERS_ID, data = covid_df)
res_all<- compareGroups( ~ . - VAERS_ID - DIED, data = covid_df)

# creating table 
table1 <- createTable(res, hide.no = 'No', show.n = FALSE) # without grouping 
table1_all <- createTable(res_all, hide.no = 'No', show.n = FALSE)

#join groups of variables by row
table1_grp <- rbind(" " = table1[1:7],
                 "Comorbidities, n(%)" = table1[8:23], 
                 "Symptoms, n(%)" =  table1[24:79])

table1_all_grp <- rbind(" " = table1_all[1:7],
                 "Comorbidities, n(%)" = table1_all[8:23], 
                 "Symptoms, n(%)" =  table1_all[24:79])

# combine two tables 
table1_comb <-cbind(table1_all_grp, table1_grp, caption = c(' ', ' ')) 
print(table1_comb, header.labels = c(all = 'Total', p.overall = 'P value'))


# export as pdf
export2pdf(table1_comb, file = 'table1.pdf', which.table="descr", nmax=TRUE, 
           header.labels=c(all = 'Total', p.overall = 'P value'), 
           caption= 'Baseline Characteristics and Clinical History of Patients with Post-COVID-19 Vaccination Adverse Reactions', 
           width=Inf, strip=FALSE, first.strip=FALSE, background="#D2D2D2",
           size=NULL,  landscape=FALSE, numcompiled=2)

```
### 3.2 Table 2 - Odd ratio table

```{r}
# Formatting 
library(rstatix)

## p-value formatting
format_p_value <- function(df){
  # indicate p_values that are less than 0.001 i.e., <0.001
  # round other p_values to 3 decimal places
  p_values <- df$P_value
  p_values <- p_format(p_values, accuracy = 0.001)
  for ( i in 1: length(p_values)){
    if (p_values[i] != '<0.001'){
      p_values[i] <- round((as.numeric(p_values[i])), digits = 3)
    }
  }
  p_values
}

p_values_odd_df <- format_p_value(odd_df)
p_values_full_df <- format_p_value(full_df)

## decimal number rounding
round_df <- function(x, digits) {
    # round all numeric variables
    # x: data frame 
    # digits: number of digits to round
    numeric_columns <- sapply(x, mode) == 'numeric'
    x[numeric_columns] <-  round(x[numeric_columns], digits)
    x
}
odd_df <- round_df(odd_df, 2)
full_df <- round_df(full_df, 2)

```

```{r adjusted for age and sex }

## factor name 
factors <- c('Hospital admission status, Yes vs No', 
'Post-vaccination disability, Yes vs No',
'Life-threatening event, Yes vs No',
'Birth defeat, Yes vs No')

for (i in 8:11){
  odd_df[i, 1] <- factors[i-7]
}

for (j in 12:26){
  odd_df[j, 1] <- str_sub(odd_df[j, 1], 1, -3)
}


## merge two confidence interval columns into one
library(tidyr)
odd_df <- unite(odd_df, CI, c(CI_2.5, CI_97.5), sep= "-", remove = FALSE)
odd_df$CI <- paste0("(", odd_df$CI, ")")
odd_df<- unite(odd_df, 'OR(95%CI)', c(OR, CI), sep = "", remove = FALSE)

log_reg_result <- data.frame(odd_df$Factors, odd_df$`OR(95%CI)`, p_values_odd_df)

## add reference line
line_age <-c('0 - 17', '1[Reference]', 'NA')
log_reg_result <-rbind(log_reg_result[1,], line_age, log_reg_result[2:84, ])
line_vac <- c('JANSSEN', '1[Reference]', 'NA')
log_reg_result <- rbind(log_reg_result[1:83,], line_vac, 
                        log_reg_result[- (1:83), ])
names(log_reg_result) <- c('Factors', 'OR(95%CI)', 'P value')

write.csv(log_reg_result, 'logit_results_1.csv')

```

```{r fully adjusted}
full_df <- unite(full_df, CI, c(CI_2.5, CI_97.5), sep= "-", remove = FALSE)
full_df$CI <- paste0("(", full_df$CI, ")")
full_df<- unite(full_df, 'OR(95%CI)', c(OR, CI), sep = "", remove = FALSE)

log_reg_result_full <- data.frame(full_df$`OR(95%CI)`, p_values_full_df)
names(log_reg_result_full) <- c('OR(95%CI)', 'P value')

## add age_group lines
line_age_group <- data.frame(c(rep(" ", 6)), c(rep(" ", 6)))
names(line_age_group) <- names(log_reg_result)[2:3]
                             
log_reg_result_full <-rbind(log_reg_result_full[1,], 
                            line_age_group,
                            log_reg_result_full[2:79,])
## add reference line
line_vac <- c('1[Reference]', 'NA')
log_reg_result_full <- rbind(log_reg_result_full[1:83,], line_vac, 
                        log_reg_result_full[- (1:83), ])

log_reg_result_full[67,] <- c('NA', 'NA')

write.csv(log_reg_result_full, 'logit_results_2.csv')
```
